---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# JointODE

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![R-CMD-check](https://github.com/signorinoy/JointODE/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/signorinoy/JointODE/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/signorinoy/JointODE/graph/badge.svg)](https://app.codecov.io/gh/signorinoy/JointODE)

<!-- badges: end -->

The **JointODE** package provides a unified framework for joint modeling of longitudinal biomarker measurements and time-to-event outcomes using ordinary differential equations (ODEs). This approach enables the simultaneous analysis of biomarker trajectories and their impact on survival outcomes.

## Model Setup

### Longitudinal Model

The observed biomarker measurements are modeled as: $$
V_{ij}=m_i(T_{ij})+b_i+\varepsilon_{ij},\quad i=1,\ldots,n,\quad j=1,\ldots,n_i
$$

where:

-   $V_{ij}$: Observed biomarker value for subject $i$ at time $T_{ij}$
-   $m_i(t)$: True underlying biomarker trajectory
-   $b_i\sim\mathcal{N}(0,\sigma_{b}^{2})$: Subject-specific random intercept
-   $\varepsilon_{ij}\sim\mathcal{N}(0,\sigma_{e}^{2})$: Measurement error

The biomarker trajectory evolution is characterized by the following second-order differential equation:

$$
\ddot{m}_i(t) = f\big(m_i(t), \dot{m}_i(t), \mathbf{X}_i(t), t\big)
$$

where $f: \mathbb{R} \times \mathbb{R} \times \mathbb{R}^p \times \mathbb{R}^+ \to \mathbb{R}$ is a smooth function modeling the biomarker acceleration as a function of its current value $m_i(t)$, velocity $\dot{m}_i(t)$, time-varying covariates $\mathbf{X}_i(t) \in \mathbb{R}^p$, and time $t$.

### Survival Model

The hazard function incorporates biomarker dynamics:

$$
\lambda_i(t) = \lambda_{0}(t)\exp\left[\mathbf{m}_i(t)^{\top}\boldsymbol{\alpha}+\mathbf{W}_i^{\top}\boldsymbol{\phi}+b_{i}\right]
$$

where:

-   $\lambda_{0}(t)$: Baseline hazard (e.g., Weibull, piecewise constant)
-   $\mathbf{m}_i(t)=\left(m_i(t), \dot{m}_i(t), \ddot{m}_i(t)\right)^{\top}$: Biomarker value and derivatives
-   $\boldsymbol{\alpha}=(\alpha_0, \alpha_1, \alpha_2)^{\top}$: Association parameters for value, velocity, and acceleration
-   $\mathbf{W}_i$: Baseline covariates with coefficients $\boldsymbol{\phi}$
-   $b_i$: Subject-specific random intercept

For detailed mathematical derivations including ODE formulation, likelihood construction, and EM algorithm specifics, see the [technical documentation](vignettes/technical-details.md).

## Installation

You can install the development version of JointODE from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("signorinoy/JointODE")
```

## Example

Here's a basic example demonstrating typical usage:

```{r example, eval=FALSE}
library(JointODE)
## basic example code
```

Using `README.Rmd` instead of `README.md` allows you to include executable R chunks:

```{r cars}
summary(cars)
```

Remember to render `README.Rmd` regularly to keep `README.md` up-to-date. Use `devtools::build_readme()` for this task.

You can also embed plots:

```{r pressure, echo = FALSE}
plot(pressure)
```

Remember to commit and push the resulting figure files to ensure they display on GitHub and CRAN.

## Code of Conduct

Please note that the JointODE project is released with a [Contributor Code of Conduct](http://gongziyang.com/JointODE/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
