---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# JointODE

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![R-CMD-check](https://github.com/ziyangg98/JointODE/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ziyangg98/JointODE/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/ziyangg98/JointODE/graph/badge.svg)](https://app.codecov.io/gh/ziyangg98/JointODE)

<!-- badges: end -->

The **JointODE** package provides a unified framework for joint modeling of longitudinal biomarker measurements and time-to-event outcomes using ordinary differential equations (ODEs). This approach enables the simultaneous analysis of biomarker trajectories and their impact on survival outcomes.

## Model Setup

### Longitudinal Model

The observed biomarker measurements are modeled as: $$
V_{ij}=m_i(T_{ij})+b_i+\varepsilon_{ij},\quad i=1,\ldots,n,\quad j=1,\ldots,n_i
$$

where:

-   $V_{ij}$: Observed biomarker value for subject $i$ at time $T_{ij}$
-   $m_i(t)$: True underlying biomarker trajectory
-   $b_i\sim\mathcal{N}(0,\sigma_{b}^{2})$: Subject-specific random intercept
-   $\varepsilon_{ij}\sim\mathcal{N}(0,\sigma_{e}^{2})$: Measurement error

The biomarker trajectory evolution is characterized by the following second-order differential equation:

$$
\ddot{m}_i(t) = f\big(m_i(t), \dot{m}_i(t), \mathbf{X}_i(t), t\big)
$$

where $f: \mathbb{R} \times \mathbb{R} \times \mathbb{R}^p \times \mathbb{R}^+ \to \mathbb{R}$ is a smooth function modeling the biomarker acceleration as a function of its current value $m_i(t)$, velocity $\dot{m}_i(t)$, time-varying covariates $\mathbf{X}_i(t) \in \mathbb{R}^p$, and time $t$.

### Survival Model

The hazard function incorporates biomarker dynamics:

$$
\lambda_i(t) = \lambda_{0}(t)\exp\left[\mathbf{m}_i(t)^{\top}\boldsymbol{\alpha}+\mathbf{W}_i^{\top}\boldsymbol{\phi}+b_{i}\right]
$$

where:

-   $\lambda_{0}(t)$: Baseline hazard (e.g., Weibull, piecewise constant)
-   $\mathbf{m}_i(t)=\left(m_i(t), \dot{m}_i(t), \ddot{m}_i(t)\right)^{\top}$: Biomarker value and derivatives
-   $\boldsymbol{\alpha}=(\alpha_0, \alpha_1, \alpha_2)^{\top}$: Association parameters for value, velocity, and acceleration
-   $\mathbf{W}_i$: Baseline covariates with coefficients $\boldsymbol{\phi}$
-   $b_i$: Subject-specific random intercept

For detailed mathematical derivations including ODE formulation, likelihood construction, and EM algorithm specifics, see the [technical documentation](http://gongziyang.com/JointODE/articles/technical-details.html).

## Installation

You can install the development version of JointODE from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("ziyangg98/JointODE")
```

## Example

Here's a basic example demonstrating typical usage:

```{r example}
library(JointODE)

# Load example dataset
data(sim)

# Fit joint ODE model
fit <- JointODE(
  longitudinal_formula = sim$formulas$longitudinal,
  longitudinal_data = sim$data$longitudinal_data,
  survival_formula = sim$formulas$survival,
  survival_data = sim$data$survival_data,
  init = sim$parameters,
  parallel = TRUE
)

# Model summary
summary(fit)

# Generate predictions
predictions <- predict(fit, times = seq(0, 10, by = 0.25))
```

## Visualization

```{r visualization, fig.width=10, fig.height=6}
library(ggplot2)
library(dplyr)
library(tidyr)

# Prepare data
ids <- c(1, 5, 10, 15)
df <- lapply(ids, \(i) {
  pred <- predictions[[i]]
  obs <- filter(sim$data$longitudinal_data, id == i)
  data.frame(
    id = i, time = pred$times,
    biomarker_est = pred$biomarker, velocity_est = pred$velocity,
    acceleration_est = pred$acceleration, survival = pred$survival
  ) %>%
    left_join(select(obs, time, v,
      biomarker_true = biomarker,
      velocity_true = velocity, acceleration_true = acceleration
    ), by = "time")
}) %>% bind_rows()

# Biomarker plot
ggplot(df, aes(x = time)) +
  geom_line(aes(y = biomarker_est, color = "Estimated")) +
  geom_line(
    aes(y = biomarker_true, color = "True"), linetype = 2, na.rm = TRUE
  ) +
  geom_point(aes(y = v, color = "Observed"), alpha = 0.3, na.rm = TRUE) +
  facet_wrap(~id) +
  theme_minimal() +
  labs(y = "Biomarker", color = "")

# Dynamics plot
df %>%
  pivot_longer(matches("(velocity|acceleration)_(est|true)"),
    names_to = c("type", "source"), names_sep = "_"
  ) %>%
  filter(!is.na(value)) %>%
  ggplot(aes(x = time, y = value, color = type, linetype = source)) +
  geom_line() +
  facet_wrap(~id, scales = "free_y") +
  theme_minimal() +
  labs(y = "Value", color = "Type", linetype = "Source")

# Survival plot
ggplot(df, aes(x = time, y = survival, color = factor(id))) +
  geom_line(linewidth = 1.2) +
  theme_minimal() +
  ylim(0, 1) +
  labs(y = "Survival Probability", color = "Subject")
```


## Code of Conduct

Please note that the JointODE project is released with a [Contributor Code of Conduct](http://gongziyang.com/JointODE/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
