---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The `JointODE` package implements joint modeling of longitudinal biomarker trajectories and survival outcomes using ordinary differential equations (ODEs). This approach captures complex biomarker dynamics while accounting for their association with event times.

## Installation

```{r setup}
# Install from GitHub
devtools::install_github("yourusername/JointODE")

# Load the package
library(JointODE)
```

## Quick Start

### Data Preparation

The package requires two data frames:

1. **Longitudinal data**: Repeated biomarker measurements
2. **Survival data**: Time-to-event outcomes

```{r data-prep}
# Example longitudinal data structure
# id: subject identifier
# time: measurement time
# value: biomarker value
head(data.long)

# Example survival data structure
# id: subject identifier
# obstime: observation time
# status: event indicator (1=event, 0=censored)
# w1, w2: baseline covariates
head(data.surv)
```

### Model Fitting

```{r model-fit}
# Fit joint ODE model
fit <- JointODE(
  data.long = data.long,
  data.surv = data.surv,
  formula.long = value ~ time,
  formula.surv = Surv(obstime, status) ~ w1 + w2,
  id = "id",
  time = "time"
)

# View results
summary(fit)
```

### Model Diagnostics

```{r diagnostics}
# Check convergence
plot(fit, type = "convergence")

# Residual diagnostics
plot(fit, type = "residuals")

# Fitted trajectories
plot(fit, type = "trajectories", subjects = 1:6)
```

## Next Steps

- See `vignette("technical-details")` for mathematical formulation
- See `vignette("data-simulation")` for generating synthetic data
- Check `?JointODE` for detailed function documentation
